International Relations are a Social Network:\\ A Latent Network Approach for Global Politics

### 2017-10-05 paper/1_intro.tex
front half

\citet{aronow:etal:2015} claim that over the period from 2010 to 2015, over sixty articles utilizing dyadic data were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization}. Most of these studies  use a generalized linear model (GLM) to estimate regression coefficients.  However, extant approaches to studying dyadic data increase the chance of faulty inferences by treating data as independent and identically distributed (iid) when observations may be highly dependent. Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters in the link function, in order to reflect the potential clustering of cases. This may work in limited situations, but is not generally effective because these palliatives do not address the fundamental data generating process which remains a threat to inference because of the interdependence of observations or measurements. It is not just the diagonals of the variance-covariance matrix this affects.

In this article, we introduce a Bayesian approach with additive and multiplicative effects for directly modeling relational data to reflect the data generating process that yields interdependencies in relational data. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads as there may be a particular actor that is more likely to send or receive events such as conflict. Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad, specifically, if a rebel group initiates a conflict with a government, the government will reciprocate that conflictual behavior. We capture these two dependencies, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. The multiplicative effects portion of the model captures dependencies that results from groups of actors clustering together or organizing into communities. Clusters and communities often result in relational data because actors possess some set of shared attributes that affect their probability of interacting with one another.

% A common example for why we observe dyads clusters in networks is due to transitivity. This is simply the principle that when considering alliance patterns we might reasonably expect that if two actors are already allied to a third that their probability of allying with one another is higher than average.

We apply this model, called the Additive and Multiplicative effects (AME) model, to five prominent studies in the international relations literature, and compare results from the current state-of-the-art approach (a GLM with robust standard errors) to those obtained with the AME framework. This latent factor approach is able to better capture \first-, \second-, and \third-order interdependencies than the standard approaches. It also produces results that are more precise, and often different, than those found in these literatures, and as such offers substantive insights, which are occluded by ignoring the interdependent nature of the relational data that characterizes many studies in the field of international relations. Finally, we show that out-of-sample predictions are substantially improved with this approach.

>>>>

\citet{aronow:etal:2015} estimate that over the period from 2010 to 2015, over sixty articles utilizing dyadic data were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization}. Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, extant approaches to studying dyadic data increase the chance of faulty inferences by treating data as independent and identically distributed (iid) when observations may be highly dependent. Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters in the link function, so as to reflect the potential clustering of cases. This may work in limited situations, but is not generally effective because these palliatives do not address the fundamental data generating process that remains a threat to inference because of the interdependence of observations or measurements. Namely, it is not just the diagonals of the variance-covariance matrix this affects.

In this article, we discuss a Bayesian approach, the Additive and Multiplicative Effects (AME) model, for directly modeling relational data to reflect the data generating process that yields interdependencies in these types of data structures \citep{hoff:2008,minhas:etal:2016:arxiv}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads as there may be a particular actor that is more likely to send or receive events such as conflict. Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad, specifically, if a rebel group initiates a conflict with a government, the government will reciprocate that conflictual behavior. We capture these two dependencies, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. The multiplicative effects portion of the model captures dependencies that results from groups of actors clustering together or organizing into communities due to meso-scopic features of networks, such as homophily and stochastic equivalence. %Clusters and communities often result in relational data because actors possess some latent set of shared attributes that affect their probability of interacting with one another.

We begin conducting a simulation study to show how AME can recover unbiased and well calibrated regression coefficients in the presence of network dependencies. Then we apply the AME model to five prominent studies in the international relations (IR) literature, and compare results from the current state-of-the-art approach (a GLM with robust standard errors) to those obtained with the AME framework. This latent factor approach is able to better capture first, second, and third-order interdependencies than the standard approaches. It also produces results that are more precise, and at times different, than those found in these literatures. As such this approach offers substantive insights, which are occluded by ignoring the interdependent nature of the relational data that characterizes many studies in the field of international relations. Finally, we show that for each replication our network based approach provides substantively more accurate out of sample predictions than the models used in the original studies. Thus the approach that we offer here is one that can be used by scholars in the field to not only generate substantive insights, but also enables us to better model the data generating process behind events of interest in international relations.

### 2017-10-05 paper/2_model.tex
front half

Relational, or dyadic, data provide measurements of how sets of actors relate to one another. Typically, pairs of actors are examined. These pairs are called \textit{dyads}. Such data can be arranged as a directed dyadic design in which the unit of analysis is some set of $n$ actors that have been paired together to form a matrix of dyads, i. e., represented as a $n \times n$ matrix, $\mathbf{Y}$, with diagonals typically undefined. The $ij^{th}$ entry defines the relationship of $i$ and $j$ and can be continuous or discrete. It can also be symmetric (known as undirected) or asymmetric (called directed). This is not the stacked dyad structure normally found in social science, but it is equivalent and presents a format in which the interdependencies are exposed and transparent. The standard statistical approach to these data uses a linear or generalized linear model (GLM) which assumes that each of the dyadic observations is conditionally independent on observed regressors, even though events and relationships between these actors is most often \textit{inter}dependent.

\begin{figure}
	\includegraphics[width=1\textwidth]{df_adj_net3.png}
	\caption{Relational data from dyadic designs to networks}
	\label{fig:ddToNet}
\end{figure}

One important dependency is nodal dependency, whereby each actor has similar interactions with many of its partners, leading to significant heterogeneity in activity levels across nodes. An important implication of this across-node heterogeneity is within-node homogeneity of ties: the values across a row, say $\{y_{ij},y_{ik},y_{il}\}$, are similar to each other because each has a common sender. If a sender $i$ is more active active in the network than other senders, it will have greater homogeneity of its linkages. More popular targets show greater similarity across linkages in a given column, $\{y_{ji},y_{ki},y_{li}\}$. Actors which are more likely to send ties in a network are also more likely to receive them. As a result the rows and columns in an adjacency matrix may be correlated. Another structural interdependence is reciprocity. This is a second-order, or dyadic dependency whereby values of $y_{ij}$ and $y_{ji}$ are correlated. If the data are simply stacked dyads in some arbitrary order, this kind of dependency is harder to visualize, but exists nonetheless.

The presence of these interdependencies in relational data complicates the practical assumption of observational independence.  Inferences drawn from models that ignore potential interdependencies between dyadic observations face a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. By ignoring these potential interdependencies, we often ignore important features of the problem under study. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

The additive part described above does not handle third-order dependency. A third-order dependency is the dependency between triads, not dyads. Third-order effects can arise from the presence of some set of shared attributes between actors that affects their probability of interacting with one another. Individuals try to close triads because this is thought to be a more stable or preferable social situation \citep{wasserman:faust:1994,zinnes:1967}.

For example, a common finding is that democracies are more likely to form trade agreements with one another. Both geographic proximity and a country's political system are examples of homophily, which captures the idea that the relationships between actors with similar characteristics in a network are likely to be stronger than nodes with different characteristics. Homophily can be used to explain the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which have a long history in international relations. A binary network where actors tend to form ties based on shared characteristics  often leads to a network with a high number of ``transitive triads'' in which  sets of actors $\{i,j,k\}$ are each linked to one another. Thus, unless we specify a list of exogenous variables that determine this prevalence of triads, the probability of $j$ and $k$ forming a tie is not independent of the ties that already exist between those actors and others.

Another third-order dependence pattern is stochastic equivalence. Actors $i,j$ are stochastically equivalent if the probability of $i$ relating to every other actor is similar to the probability for $j$. Equivalence implies the existence of groups of nodes in a network with similar relational patterns.  Recent work has found stochastic equivalence to be important \citep{manger:etal:2012}. Latent variables can account for third-order dependence patterns within the context of the SRRM. These models assume that relationships between nodes are mediated by a small number ($K$) of node-specific unobserved latent variables.

@@

The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science.  It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence.  We do not know whether interdependence dominates international relations, but as noted by Thoreau in reference to rumors that dairymen on strike were watering down milk: ``\ldots some circumstantial evidence is very strong, as when you find a trout in the milk.''  At a very minimum, it is necessary to examine whether there is interdependence since it challenges substantive arguments as well as statistical modeling \citep{snijders:2011}. Ignoring the assumption of independence is a widely-noted shortcoming of many studies, but it continues to be the standard practice in much scholarship. Fortunately, it is easy to empirically examine dependencies in relational data.

The additive and multiplicative effects model (AME) can help to eliminate this color-blindness to interdependence. It was introduced in \cite{hoff:2008}; the open-source computer package associated with this approach is \cite{amenpkg}. We describe it below.

Parameter interpretation is straightforward because it has a simple regression basis as shown in Equation~\ref{eqn:ame} which portrays directed matrix \:

@@

With this framework, it is straightforward to model dyadic observations as conditionally independent given $\bm\theta$, where $\bm\theta$ depends on the the unobserved random effects modeled to account for the potential \first, \second, and \third-order dependencies. In Equation~\ref{eqn:srmCov}, $a_{i} + b_{j}  + \epsilon_{ij}$ represent the additive random effects in this framework and account for sender, receiver, and within-dyad dependence. The multiplicative effects, $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}$, capture higher-order dependence patterns in $\bm\theta$ after accounting for any known covariate information. A Bayesian procedure in which parameters are iteratively updated using a Gibbs sampler is available in the \pkg{amen} $\sf{R}$ package to estimate this type of generalized linear mixed effects model from Gaussian, binary, ordinal, and other relational data types.

\subsection{\textbf{Social Relations Model: Additive Part of AME}}

\citet{warner:etal:1979} developed a social relations model (SRM), using ANOVA decomposition, that focuses on such dependencies. SRM provides the error structure for the additive effects component of the AME framework. SRM decomposes the variance of dyadic observations in terms of a) heterogeneity across row means (often called out-degree), b) column means (or, in-degree), c) correlation between row and column means, and d) correlations within dyads.  \citet{wong:1982,li:loken:2002} provide a random effects representation of the SRM:

\begin{align}
\begin{aligned}
	y_{ij} &= \mu + e_{ij} \\
	e_{ij} &= a_{i} + b_{j} + \epsilon_{ij} \\
	\{ (a_{1}, b_{1}), \ldots, (a_{n}, b_{n}) \} &\simiid N(0,\Sigma_{ab}) \\ 
	\{ (\epsilon_{ij}, \epsilon_{ji}) : \; i \neq j\} &\simiid N(0,\Sigma_{\epsilon}), \text{ where } \\
	\Sigma_{ab} = \begin{pmatrix} \sigma_{a}^{2} & \sigma_{ab} \\ \sigma_{ab} & \sigma_{b}^2   \end{pmatrix} \;\;\;\;\; &\Sigma_{\epsilon} = \sigma_{\epsilon}^{2} \begin{pmatrix} 1 & \rho \\ \rho & 1  \end{pmatrix},
\label{eqn:srmCov}
\end{aligned}
\end{align}

wherein $\mu$ is the density of a network, and $e_{ij}$ represents residual variation in each cell (or dyad). The residual variation has three separate parts: a)   row/sender effects ($a_{i}$), b)  column/receiver effects ($b_{j}$), and c) within-dyad effects ($\epsilon_{ij}$). The row and column effects are modeled jointly to account for correlation in how active an actor is in sending and receiving ties. Heterogeneity in the row and column means is modeled by $\sigma_{a}^{2}$ and $\sigma_{b}^{2}$, respectively.  $\sigma_{ab}$ describes whether actors which send a lot of ties also receive a lot. Second-order dependencies are described by $\sigma_{\epsilon}^{2}$ and a reciprocity, parameter $\rho$.

The SRM covariance structure described in Equation~\ref{eqn:srmCov} can be incorporated into the systematic component of a GLM framework to produce the social relations regression model (SRRM): $\bm\beta^{\top} \mathbf{X}_{ij} + a_{i} + b_{j} + \epsilon_{ij}$, where $ \bm\beta^{\top} \mathbf{X}_{ij}$ accommodates the inclusion of dyadic, sender, and receiver covariates \citep{hoff:2005}. This is a regression framework with additive, random effects, but it models variances and covariances often seen in relational data. Furthermore, this accommodates a diversity of outcome distributions. In the case of binary data, this can be done by utilizing a random effects logistic (or probit) latent variable representation.

\subsection{\textbf{Latent Factor Model: Multiplicative Part of AME}}

The latent factor model (LFM) expands the SSRM to include an additional term, $y_{ij}$, $\alpha(u_{i}, u_{j})$, that captures latent third-order characteristics of a network---where $u_{i}$ and $u_{j}$ are node-specific latent variables for symmetric data. General definitions for how $\alpha(u_{i}, u_{j})$ is defined for these latent variable models are shown in Equation~\ref{eqn:latAlpha}.

\begin{align}
\begin{aligned}
	\alpha(&\textbf{u}_{i}, \textbf{u}_{j}) = \textbf{u}_{i}^{\top} \Lambda \textbf{u}_{j} \\
	&\textbf{u}_{i} \in \mathbb{R}^{K}, \; i \in \{1, \ldots, n \} \text{, where}\\
	 &\Lambda \equiv K \times K \text{ diagonal matrix}
\label{eqn:latAlpha}
\end{aligned}
\end{align}

Many observed  networks exhibit varying degrees of stochastic equivalence and homophily.  In the latent factor model, each actor has an unobserved vector of characteristics, $\textbf{u}_{i} = \{u_{i,1}, \ldots, u_{i,K} \}$, describing each node's behavior in the network. The probability of a tie between $i$ and $j$ depends on the extent to which $\textbf{u}_{i}$ and $\textbf{u}_{j}$ are ``similar'' and on whether the entries of $\Lambda$ are greater or less than zero. Similarity in the latent factors, $\textbf{u}_{i} \approx \textbf{u}_{j}$, corresponds to stochastic equivalence; the eigenvalue determines whether the network exhibits positive or negative homophily. Accordingly, this approach can represent both homophily and stochastic equivalence.

In directed (asymmetric) networks there are two sets of latent positions, denoted $\textbf{u}$ and $\textbf{v}$, representing the sending and receiving positions separately.  Thus, actors in the network have a vector of latent characteristics describing their behavior as a sender, denoted by $\textbf{u}$, and as a receiver, $\textbf{v}$: $\textbf{u}_{i}, \textbf{v}_{j} \in \mathbb{R}^{K}$. This can alter the probability of an interaction between $ij$ additively: $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}$, where $\textbf{D}$ is a $K \times K$ diagonal matrix.

The LFM captures a wide range of interdependence patterns that arise in relational data such as often studied in international relations.

>>>>

In thinking about dyadic data, scholars in the field begin by structuring it as a set of dyadic observations stacked on top of one another. Each observation is assumed to be independent of the others. Thus, for example, a conflict sent from actor the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. As a result, the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others.

\input{dd_to_adjmat}
\FloatBarrier

To start to move away from this assumption and better understand the dependencies that emerge in relational data it is helpful to shift towards structuring dyadic data in the form of an adjacency matrix as shown in the top right of Table~\ref{tab:netDesign}. Here rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the action that was sent by an actor on the row to one designated in the column. Thus $y_{ij}$ designates an action, such as a conflictual event or trade flows, that is sent from actor $i$ to actor $j$.

Using the structure of an adjacency matrix we can visualize the types of first and second order dependencies that complicate the analysis of relational data in traditional GLMs. Figure~\ref{fig:adjMatDeps} clarifies the types of dependencies that can manifest in these types of data structures. The adjacency matrix on the top left highlights a particular row of an adjacency matrix, to illustrate that values across a particular row of an adjacency matrix may be more similar to each other than other values in the adjacency matrix because each of these values has a common sender. Homogeneity in interactions involving a common sender also manifest heterogeneity in how active actors are across the network when compared to each other. Thus in most relational datasets (e.g., trade flows, conflict, participation in international organizations, even networks derived from Twitter or Facebook) we often find that there are some actors who are much more active than others \citep{barabasi:reka:1999}. Unless one is able to develop a model that can account for the variety of explanations that may play a role in determining why a particular actor may be more active than others, parameter estimates from standard statistical models will be biased.

\begin{figure}
	\begin{tabular}{cc}
		\texttt{Sender heterogeneity} & \texttt{Receiver Heterogeneity} \\
		\includegraphics[width=.45\textwidth]{adjRowDep.png} & \includegraphics[width=.45\textwidth]{adjColDep.png} \\
		\texttt{Sender-Receiver Covariance} & \texttt{Reciprocity} \\
		\includegraphics[width=.45\textwidth]{adjRowColCovar.png} & \includegraphics[width=.45\textwidth]{adjRecip.png} \\
	\end{tabular}
	\caption{Nodal and dyadic dependencies in relational data.}
	\label{fig:adjMatDeps}
\end{figure}

For similar reasons one also needs to take into account that there is a shared dependence between dyadic observations that share a common receiver. The bottom-left panel, illustrates that sender and receiver type dependencies can also blend together. Specifically, as actors who are more likely to send ties in a network tend to also be more likely to receive them. As a result, the rows and columns in an adjacency matrix are often correlated. For example, consider that trade flows both from and to many wealthy, developed countries. The bottom-right panel, highlights a second order dependence, specifically, reciprocity. This is a dependency occurring within dyads involving the same actors whereby values of $y_{ij}$ and $y_{ji}$ are correlated. The concept of reciprocity has deep roots in the study of relations between states \citep{richardson:1960,keohane:1989}. The purpose of highlighting each of these dependencies through the set of panels in Figure~\ref{fig:adjMatDeps} is to more easily visualize how each of these dependencies manifest in relational data. Alternatively, when we simply simply stack dyads in some arbitrary order, these dependencies are easy to ignore.

The dependencies discussed so far involve at most dependence between two actors. For most relational data, however, dependencies do not simply manifest at this level. More often we find significant evidence of higher order structures that result from dependencies between multiple groups of actors. These dependencies arise because there may be a or some set of latent attributes between actors that affects their probability of interacting with one another \citep{wasserman:faust:1994,zinnes:1967}. In Figure~\ref{fig:thirdDeps} we provide a visualization of a hypothetical relational dataset. Here the nodes designate actors and edges between the nodes indicate that an interaction between the two took place. Each node is colored by the latent group to which they belong.

\begin{figure}[ht]
	\includegraphics[width=.7\textwidth]{stochEquiv.pdf}
	\caption{Visualization of network with meso-scopic features.}
	\label{fig:thirdDeps}
\end{figure}

Clear from the visualization is that the actors belonging to the same group have a higher likelihood of having an interaction with each other than those from the other group. A prominent example of a network with this type of structure was found by \citet{adamic:glance:2005}, who visualized the ways in which right and left leaning political blogs linked to one another in the 2004 United States Election \citep{adamic:glance:2005}. They found that the degree of interaction between right and left leaning blogs was minimal, and that most of these blogs simply linked to those of their own ilk. This showcases the types of higher order dependencies that can emerge in relational data. First, the fact that interactions was determined by a shared attribute, in this case political ideology, is an example of homophily. Homophily can be used to explain the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. The other major type of meso-scopic feature that emerges in relational data is community structure, which is often formalized through the concept of stochastic equivalence \citep{anderson:etal:1992}. This concept simply refers to the idea that groups of nodes that act similarly in the network are stochastically equivalent. In the example we have laid out above each of the left leaning blogs would be considered stochastically equivalent to one another.

The major implication of the presence of homophily, stochastic equivalence, and the other dependencies we discussed above are that they complicate the practical assumption of observational independence. Inferences drawn from models that ignore potential interdependencies between dyadic observations face a number of well-known challenges: a) biased estimates of the effect of independent variables, b) uncalibrated confidence intervals, and c) poor predictive performance. By ignoring these potential interdependencies, we often ignore important features of the problem under study. The study of international relations is founded on the relations among actors. Why ignore the interdependencies that led to the study of IR in the first place?

% Thus, unless we specify a list of exogenous variables that determine this prevalence of triads, the probability of $j$ and $k$ forming a tie is not independent of the ties that already exist between those actors and others.

@@

The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science. It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence.  We do not know whether interdependence dominates international relations, but as noted by Thoreau in reference to rumors that dairymen on strike were watering down milk: ``\ldots some circumstantial evidence is very strong, as when you find a trout in the milk.'' At a very minimum, it is necessary to examine whether there is interdependence since it challenges substantive arguments as well as statistical modeling \citep{snijders:2011}.

Further parameter interpretation in the AME framework is straightforward because it has a simple regression basis as shown in Equation~\ref{eqn:ame} which portrays directed matrix \:

@@

With this framework, it is straightforward to model dyadic observations as conditionally independent given $\bm\theta$, where $\bm\theta$ depends on the the unobserved random effects modeled to account for the potential \first, \second, and \third-order dependencies. In Equation~\ref{eqn:srmCov}, $a_{i} + b_{j}  + \epsilon_{ij}$ represent the additive random effects in this framework and account for sender, receiver, and within-dyad dependence. The multiplicative effects, $\textbf{u}_{i}^{\top} \textbf{D} \textbf{v}_{j}$, capture higher-order dependence patterns in $\bm\theta$ after accounting for any known covariate information.\footnote{This procedure is described in detail in \citet{hoff:2008} and \citet{minhas:etal:2016:arxiv}. A Bayesian procedure in which parameters are iteratively updated using a Gibbs sampler is described in the appendix.}
